---
title: "QC of metabolite data"
author: "David Hughes"
date: "1/9/2022"
output: 
  pdf_document:
    number_sections: true
---

This Rmd file will identify for each metabolite and metabolite pair (fasting & postprandial), outliers in the distribution and turn them into NAs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 4, echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(quantreg)
# library(patchwork)
```


## Load the data

```{r}
#########################
### LOAD Data
#########################
### parameter file
pfile = "parameters/pfile.txt"
par_data = read.table(pfile, header = FALSE, sep = "=", as.is = TRUE)

## load study R data object
cat(paste0("Load the study data Rdata file\n"))
w = grep("data_dir", par_data[,1])
f = paste0(par_data[w,2], "processed/my_working_QCd_study_data.Rdata" )
load(f)

## Define mydata: the working data object
cat(paste0("Define mydata working data frame\n"))
mydata = study_data$working_data

```


# 'mydata' is my working study data set

## Perform metabolite quality control

1. ensure zero values are removed (turned to NA)
2. for each feature identify values that are 10 IQR distances from the median and remove (turn to NA)
3. In a metabolite paired, fasting - postprandial framework, estimate a simple euclidean delta and identify values that are 5 IQR distances from the median and remove (turn to NA)

## Define a vector of all of the trait names and a vector of all of the metabolite IDs

```{r}
metids = study_data$metids
traits = study_data$traits
```

## Remove outlier samples observed in at each trait.

```{r}
## Load function
source("functions/id_outliers.R")
##
outlier_count = c()
##
for(trait in traits){
  ## turn 0 values into NA
  zerovals = which(mydata[, trait] == 0)
  if(length(zerovals)>0){
    mydata[zerovals, trait] = NA
  }
  ## Identify outliers
  outliers = id_outliers( mydata[, trait], 
                          outlier_method = "iqr", 
                          outlier_cutoff = 10 )
  ## Outlier Counter
  outlier_count[trait] = length(outliers)
  
  ## turn outliers into NA
  if(length(outliers)>0){
    mydata[outliers, trait] = NA
  }
}

#####
```

## How many extreme outliers did we identify and turn into NA?

```{r}
summary(outlier_count)
```

```{r}
table(outlier_count)
```
## traits ranked by the number of extreme outliers identified and turned into NA.

```{r}
sort(outlier_count, decreasing = TRUE)[1:40]
```


## Remove outlier samples at trait pairs (fasting -v- postprandial)

```{r}
paired_outlier_count = c()
###
for(metid in metids){
  ## define the fasting and postprandial states of the metid
  fasting = paste0(metid, "_f")
  postprandial = paste0(metid, "_p")
  
  ## Estimate delta
  xydelta = mydata[, postprandial] - mydata[, fasting]
  
  ## Identify the extreme outliers
  xydelta_outliers = id_outliers( xydelta,
                                outlier_method = "iqr", 
                                outlier_cutoff = 5)
  
  ## Outlier Counter
  paired_outlier_count[metid] = length(xydelta_outliers)
  
  ## Turn any outliers into NA
  if(length(xydelta_outliers)>0){
     mydata[xydelta_outliers, fasting] = NA
     mydata[xydelta_outliers, postprandial] = NA
    }
    
}

```


## How many extreme outliers did we identify and turn into NA?

```{r}
summary(paired_outlier_count)
```

```{r}
table(paired_outlier_count)
```
## traits ranked by the number of extreme outliers identified and turned into NA.

```{r}
sort(paired_outlier_count, decreasing = TRUE)[1:40]
```


```{r}
study_data$working_data = mydata
```


## Write study_data R object to file

```{r}
w = grep("data_dir", par_data[,1])
f = paste0(par_data[w,2], "processed/my_working_QCd_study_data_v1.Rdata" )
save(study_data, file = f)
```



