---
title: "NEO BMI-Metabolite covariables correlations"
author: "David Hughes"
date: "2/6/2021"
output: 
  pdf_document:
    number_sections: true
---

This report will evaluate the correlation between BMI & BMI-PGS and all study covariables


```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = FALSE, error = FALSE, warning = FALSE, message = FALSE)

source("LOAD.R")

```


## Correlation structure among these covariables

## How many unique covariables are there ?

## estimate covariables cormat

```{r, fig.width = 12, fig.height = 6}
library(NILC)

########################
## define working data set
########################
wdata = mydata[, covariables ]

########################
## Estimate correlations
########################
## NILC function
# cormat = Test_DF_Correlations( wdata )
## modified function
source("functions/cormat_nf.R")
cormat = cormat_nf( wdata )

## Add Names
rownames(cormat$RhoMat) = colnames(wdata)
colnames(cormat$RhoMat) = colnames(wdata)

### abs(R)
dmat = as.dist( 1-abs(cormat$RhoMat) )

## R^2 distances
# dmat = as.dist( 1 - (cormat$RhoMat)^2 )

## estimate tree
nj = hclust(dmat, method = "complete")
```

### run a tree cut 

```{r}
## Run Tree cut
cat(paste0("tree cut height 0.5\n") )
k = cutree(nj, h = 0.5)
length( table(k) ) 

cat(paste0("\ntree cut height 0.8\n") )
k = cutree(nj, h = 0.8)
length( table(k) ) 

```

```{r}
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "figures/covariable_dendrogram.pdf")
####
pdf(f, width = 15, height = 6)
plot(nj, hang = -1, xlab = "cut height 0.5, Me = 54; cut height 0.8, Me = 37", sub = "", title = "covariable heirchical clustering")
abline(h = 0.5, col = "red", lwd = 2)
abline(h = 0.8, col = "blue", lwd = 2)
dev.off()

```


```{r, fig.width = 15, fig.height = 6}
plot(nj, hang = -1, xlab = "cut height 0.5, Me = 54; cut height 0.8, Me = 37", sub = "", title = "covariable heirchical clustering")
abline(h = 0.5, col = "red", lwd = 2)
abline(h = 0.8, col = "blue", lwd = 2)
```


# Correlations with possible confounders

## Correlation among possible confounders and BMI, BMI-GRS, and d.hat


```{r, echo = FALSE}
################################
##  Weighted NEO sample
################################
wNEO_BMI = covariate_lms(datain = mydata, dependent = "bmi", 
                        covariates = covariables[-c(5)], 
                        weights = mydata$pweight)
wNEO_BMI$sample = "wNEO"

wNEO_PGS = covariate_lms(datain = mydata, dependent = "Yengo_Weighted_GRS", 
                        covariates = covariables[-c(5)],
                        weights = mydata$pweight)
wNEO_PGS$sample = "wNEO"

################################
##   NEO sample
################################
NEO_BMI = covariate_lms(datain = mydata, dependent = "bmi", 
                        covariates = covariables[-c(5)] )
NEO_BMI$sample = "NEO"

NEO_PGS = covariate_lms(datain = mydata, dependent = "Yengo_Weighted_GRS", 
                        covariates = covariables[-c(5)] )
NEO_PGS$sample = "NEO"

################################
##  Leiderdrop sub-sample
################################
w = which( mydata$subpop == "Leiderdorp")
Leiderdorp_BMI = covariate_lms(datain = mydata[w, ], 
                               dependent = "bmi", 
                               covariates = covariables[-c(1,5)])
Leiderdorp_BMI$sample = "Leiderdorp"

Leiderdorp_PGS = covariate_lms(datain = mydata[w, ], 
                               dependent = "Yengo_Weighted_GRS", 
                               covariates = covariables[-c(1,5)])
Leiderdorp_PGS$sample = "Leiderdorp"

################################
##  Leiden
################################
w = which( mydata$subpop == "Leiden")
Leiden_BMI = covariate_lms(datain = mydata[w, ], 
                          dependent = "bmi", 
                          covariates = covariables[-c(1,5)])
Leiden_BMI$sample = "Leiden"

Leiden_PGS = covariate_lms(datain = mydata[w, ], 
                             dependent = "Yengo_Weighted_GRS", 
                             covariates = covariables[-c(1,5)])
Leiden_PGS$sample = "Leiden"


#######################
## Combine the data
#######################
confounders_cor = rbind(wNEO_BMI, wNEO_PGS, 
                        Leiderdorp_BMI, Leiderdorp_PGS,
                        NEO_BMI, NEO_PGS,
                        Leiden_BMI, Leiden_PGS)

```

```{r}

w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "tables/Dec13_2022_BMI_GRS_covariate_cors.txt")
write.table(confounders_cor, file = f, 
             row.names = TRUE, col.names = TRUE, sep = "\t", quote = FALSE)

```


```{r}
### Read in data written to file
# confounders_cor = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```



## what is BMI correlated with in the wNEO sample?

```{r}
wNEO_BMI = confounders_cor %>% filter(sample == "wNEO", dependent == "bmi") %>% arrange(pvalue)

## add categories to wNEO_BMI
m = match(wNEO_BMI$covariate, study_data$covariate_anno$new_names)
wNEO_BMI$category = study_data$covariate_anno$category[m]
##
cat(paste0("nominally associated count\n"))
w = which( wNEO_BMI$pvalue <= 0.05  )
length(w)
##
cat(paste0("\nassociated count\n"))
q = which( wNEO_BMI$pvalue <= ( 0.05/54 ) )
length(q)
##
cat(paste0("\nEta-squared summary across all tested variables\n"))
summary(wNEO_BMI$etasq)
###
cat(paste0("\nvariables associated with BMI\n"))
wNEO_BMI$covariate[q]
```

```{r}

```



```{r}
temp = wNEO_BMI[q,] %>% arrange(desc(etasq))
temp$covariate = factor(temp$covariate , levels = temp$covariate[nrow(temp):1] )
### ORDER Category by Mean etasq
od = temp %>% group_by(category) %>% summarize(mean = mean(etasq)) %>% arrange(desc(mean))
o = od$category
temp$category = factor(temp$category, levels = o)
##
od %>% kable() %>% kableExtra::kable_classic()
```

```{r}
wNEO_BMI %>% filter(category %in% c("smoking", "education", "income") ) %>% 
  kable() %>% 
  kableExtra::kable_classic()
```




```{r, fig.width = 8, fig.height = 15}
### Color Scheme
pcol = c( RColorBrewer::brewer.pal(11, "Set1"), wesanderson::wes_palette(name = "Rushmore1", n = 5) )
###
PlotA = temp %>% ggplot( aes(x = covariate, y = etasq) ) +
  geom_bar(stat = "identity", aes(fill = category) ) +
  scale_fill_manual(values = pcol) +
  coord_flip() +
  labs(y = "variance explained", x = "covariable") +
  theme_bw()

PlotA
```


## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/wNEO_BMI_covariable_etasq.pdf")
pdf(f, width = 8, height = 15)
PlotA
dev.off()


```


## what is PGS correlated with in the wNEO sample?

```{r}
wNEO_PGS = confounders_cor %>% filter(sample == "wNEO", dependent == "Yengo_Weighted_GRS") %>% arrange(pvalue)
## add categories to wNEO_BMI
m = match(wNEO_PGS$covariate, study_data$covariate_anno$new_names)
wNEO_PGS$category = study_data$covariate_anno$category[m]
###
cat(paste0("nominally associated count\n"))
w = which( wNEO_PGS$pvalue <= 0.05  )
length(w)
##
cat(paste0("\nassociated count\n"))
q = which( wNEO_PGS$pvalue <= ( 0.05/54 ) )
length(q)
##
cat(paste0("\nEta-squared summary across all tested variables\n"))
summary(wNEO_PGS$etasq)
###
cat(paste0("\nvariables associated with PGS\n"))
wNEO_PGS$covariate[q]
```



```{r}
temp = wNEO_PGS[q,] %>% arrange(desc(etasq))
temp$covariate = factor(temp$covariate , levels = temp$covariate[nrow(temp):1] )
### ORDER Category by Mean etasq
od = temp %>% group_by(category) %>% summarize(mean = mean(etasq)) %>% arrange(desc(mean))
o = od$category
temp$category = factor(temp$category, levels = o)
##
od %>% 
  kable() %>% 
  kableExtra::kable_classic()
```


```{r, fig.width = 8, fig.height = 6}
pcol = RColorBrewer::brewer.pal(8, "Set1")
###
PlotA = temp %>% ggplot( aes(x = covariate, y = etasq) ) +
  geom_bar(stat = "identity", aes(fill = category)) + 
  scale_fill_manual(values = pcol) +
  coord_flip() +
  labs(y = "variance explained", x = "covariable") +
  theme_bw()

PlotA
```


## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/wNEO_PGS_covariable_etasq.pdf")
pdf(f, width = 8, height = 6)
PlotA
dev.off()


```



```{r edit_confounder_cors, echo = FALSE}
########################################################################
### Set those P values that are 0, 
###   i.e. those that could not be estimated to the smallest value observed
########################################################################
w = which( confounders_cor$pvalue == 0)
if(length(w)>0){
  confounders_cor$pvalue[w] = min( confounders_cor$pvalue[-w])  
}

####################################
## Set those traits with a P larger 
## than a BF Pvalue to 0
####################################
w = which(confounders_cor$pvalue > 0.05/54 ) ## 54 determined by tree cut above
if(length(w)>0){ confounders_cor$pvalue[w] = 0 }

##############################
## Redefine dependent labels
##############################
w = which(confounders_cor$dependent == "bmi")
confounders_cor$dependent[w] = "BMI"

w = which(confounders_cor$dependent == "Yengo_Weighted_GRS")
confounders_cor$dependent[w] = "PGS"

```


```{r confounder_raster_bmi, fig.width = 15, fig.height = 6}
##################
## Plot colors
##################
# pcol =  wesanderson::wes_palette("Darjeeling2", 5 , type = "discrete")
pcol =  wesanderson::wes_palette("Zissou1", 5, type = "discrete")
# pcol =  wesanderson::wes_palette("BottleRocket2", 15, type = "continuous")


### Order by BMI wNEO etasq
w = unlist( confounders_cor %>% filter(dependent == "BMI" & sample == "wNEO") %>% summarize(order(etasq, decreasing = TRUE)) )
trait_order = confounders_cor$covariate[w]
confounders_cor$covariate = factor( confounders_cor$covariate, levels = trait_order)

## Define Sample levels
confounders_cor$sample = factor(confounders_cor$sample, levels = c("wNEO","Leiderdorp","NEO","Leiden"))

###
PlotA = confounders_cor %>% ggplot( aes(y = dependent, x = covariate) ) +
  geom_tile(aes(fill = etasq)) +
  scale_fill_gradient2(mid = "white" , low = pcol[4], high = pcol[1], midpoint = 0.002   ) +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_point(aes(color = -log10(pvalue) , size = -log10(pvalue) ) , shape = 19 ) +
  scale_size(range = c(1, 8), breaks = c(3, 5, 10, 20, 50, 200), labels =c("3", "5", "10", "20", "50", "200") ) +
  scale_color_gradient(low = "pink", high = "red") + 
  labs(color = "-log10(P)", size = "-log10(P)", y = "") +
  theme(legend.position="bottom") +
  facet_wrap(.~sample, nrow = 4)


PlotA


```

## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/PGS_covariate_EtaSq.pdf")
pdf(f, width = 9, height = 6)
PlotA
dev.off()


```

```{r confounder_raster_bmi, fig.width = 10, fig.height = 20}
##################
## Plot colors
##################
# pcol =  wesanderson::wes_palette("Darjeeling2", 5 , type = "discrete")
pcol =  wesanderson::wes_palette("Zissou1", 5, type = "discrete")
# pcol =  wesanderson::wes_palette("BottleRocket2", 15, type = "continuous")


### Order by BMI wNEO etasq
w = unlist( confounders_cor %>% filter(dependent == "BMI" & sample == "wNEO") %>% summarize(order(etasq, decreasing = TRUE)) )
trait_order = confounders_cor$covariate[w]
confounders_cor$covariate = factor( confounders_cor$covariate, levels = trait_order[length(trait_order):1])

## Define Sample levels
confounders_cor$sample = factor(confounders_cor$sample, levels = c("wNEO","Leiderdorp","NEO","Leiden"))

###
PlotA = confounders_cor %>% ggplot( aes(y = dependent, x = covariate) ) +
  geom_tile(aes(fill = etasq)) +
  scale_fill_gradient2(mid = "white" , low = pcol[4], high = pcol[1], midpoint = 0.002   ) +
  theme(axis.text.x = element_text(angle = 90)) +
  # geom_point(aes(color = -log10(pvalue) , size = -log10(pvalue) ) , shape = 19 ) +
  geom_point(aes( size = -log10(pvalue) ) , shape = 19, color = "red" ) +
  scale_size(range = c(1, 8), breaks = c(3, 5, 10, 20, 50, 200), labels =c("3", "5", "10", "20", "50", "200") ) +
  scale_color_gradient(low = "pink", high = "red") + 
  geom_text(aes(label = round(etasq*100, d = 2)), nudge_y = 0.01, size = 4, color = "grey20") +
  labs(color = "-log10(P)", size = "-log10(P)", y = "") +
  theme(legend.position="bottom") +
  coord_flip() +
  facet_wrap(.~sample, ncol = 4)


PlotA


```


```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/PGS_covariate_EtaSq_flip.pdf")
pdf(f, width = 10, height = 20)
PlotA
dev.off()


```


## Plot only those covariates associated with PGS in wNEO

```{r confounder_raster_bmi_v2, fig.width = 8, fig.height = 5}
temp = confounders_cor %>% 
  filter( sample == "wNEO" & dependent == "PGS" & pvalue < 0.05/54 & pvalue != 0 ) 

covars2keep = temp$covariate

confounders_cor_v2 = confounders_cor %>% filter(covariate %in% covars2keep & dependent == "PGS")
#############
### Order by PGS wNEO etasq
w = unlist( confounders_cor_v2 %>% filter(dependent == "PGS" & sample == "wNEO") %>% summarize(order(pvalue, decreasing = FALSE)) )
trait_order = as.character( confounders_cor_v2$covariate[w] )
confounders_cor_v2$covariate = factor( confounders_cor_v2$covariate, levels = trait_order)

#############
pcol =  wesanderson::wes_palette("Zissou1", 5, type = "discrete")

PlotA = confounders_cor_v2 %>% ggplot( aes(y = dependent, x = covariate) ) +
  geom_tile(aes(fill = etasq)) +
  scale_fill_gradient2(mid = "white" , low = pcol[4], high = pcol[1], midpoint = 0.002   ) +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_point(aes(color = -log10(pvalue) , size = -log10(pvalue) ) , shape = 19 ) +
  scale_size(range = c(1, 8), breaks = c(3, 5, 10, 20, 50, 200), labels =c("3", "5", "10", "20", "50", "200") ) +
  scale_color_gradient(low = "pink", high = "red") + 
  labs(color = "-log10(P)", size = "-log10(P)", y = "") +
  theme(legend.position="bottom") +
  facet_wrap(.~sample, nrow = 4)


PlotA

```

```{r, fig.width = 10, fig.height = 8}
temp = confounders_cor %>% 
  filter( sample == "wNEO" & dependent == "PGS" & pvalue < 0.05/54 & pvalue != 0 ) 

covars2keep = temp$covariate

confounders_cor_v2 = confounders_cor %>% filter(covariate %in% covars2keep & dependent == "PGS")
#############
### Order by PGS wNEO etasq
w = unlist( confounders_cor_v2 %>% filter(dependent == "PGS" & sample == "wNEO") %>% summarize(order(pvalue, decreasing = FALSE)) )
trait_order = as.character( confounders_cor_v2$covariate[w] )
confounders_cor_v2$covariate = factor( confounders_cor_v2$covariate, levels = trait_order[length(trait_order):1] )

#############
pcol =  wesanderson::wes_palette("Zissou1", 5, type = "discrete")

PlotA = confounders_cor_v2 %>% ggplot( aes(y = dependent, x = covariate) ) +
  geom_tile(aes(fill = etasq)) +
  scale_fill_gradient2(mid = "white" , low = pcol[4], high = pcol[1], midpoint = 0.002   ) +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_point(aes(color = -log10(pvalue) , size = -log10(pvalue) ) , shape = 19 ) +
  scale_size(range = c(1, 8), breaks = c(3, 5, 10, 20, 50, 200), labels =c("3", "5", "10", "20", "50", "200") ) +
  scale_color_gradient(low = "pink", high = "red") + 
  geom_text(aes(label = round(etasq*100, d = 2)), nudge_y = 0.01, size = 5, color = "black", alpha = 0.8) +
  labs(color = "-log10(P)", size = "-log10(P)", y = "", x = "covariables") +
  theme(legend.position="bottom") +
  coord_flip() +
  facet_wrap(.~sample, ncol = 4)


PlotA
```


## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/PGS_covariate_EtaSq_v2.pdf")
pdf(f, width = 8, height = 8)
PlotA
dev.off()


```

## PC3 and sub_population

```{r}
myest = c()

for(pc in c("PC1","PC2","PC3","PC4")){
  form = formula( paste0(pc, "~ subpop") )
  fit = lm(form, data = mydata)
  p = summary(fit)$coef[2,4]
  a = anova(fit)
  etsq = a[1,2]/sum(a[,2])
  myest = rbind(myest, c(etsq, p) )
}

rownames(myest) = c("PC1","PC2","PC3","PC4")
colnames(myest) = c("etasq","P")
myest
```
```{r}
fit0 = lm(Yengo_Weighted_GRS ~  PC3, data = mydata, weights = mydata$pweight)
summary(fit0)$coef
anova(fit0)

fit = lm(Yengo_Weighted_GRS ~ subpop + PC3, data = mydata, weights = mydata$pweight)
anova(fit)
```



```{r}
mytext1 = paste0("% variance explained in PC3 by subpop")
mytext2 = paste0("etasq*100 = ", round(myest[3,1]*100, d = 2), "%; P = ", signif(myest[3,2], d = 3) )

p1 = mydata %>% ggplot(aes(x = PC3, y = PC4)) +
  geom_point(aes(color = subpop), alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()


p2 = p1 + annotate(geom = "text", x = 0.025, y = -0.01, label = mytext1) +
  annotate(geom = "text", x = 0.025, y = -0.0125, label = mytext2) 

p2

```


## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/PC3_subpop_structure.pdf")
pdf(f, width = 8, height = 5)
p2
dev.off()


```



## Correlation among all possible confounders and all metabolites

```{r}

vars_2_test = c( as.character( unique(temp$covariate) ), "edu_higher", "alcohol" )
###########################
metabo_confounder_cors = c()
for(trait in traits){
  # cat(paste0(trait, "\n"))
  for(v2t in vars_2_test){
    # cat(paste0(v2t, "\n"))
    ## define new data frame 
    df = na.omit( mydata[, c(trait, v2t, "pweight")] )
    ## test for no variation
    if( length( unique( df[,2] ) ) == 1){
      test = c(trait, v2t, nrow(df), NA, NA)
      names(test) = c("dependent","covariate","sample_size","etasq","pvalue")
    } else {
      ##
      test = covariate_lms(datain = df, 
                       dependent = trait, 
                       covariates = v2t, 
                       weights = df$pweight) 
    
    }
    metabo_confounder_cors = rbind(metabo_confounder_cors, test)
  }
}
  
metabo_confounder_cors = data.frame(metabo_confounder_cors)
for(i in 3:5){
  metabo_confounder_cors[,i] = as.numeric(metabo_confounder_cors[,i])
}
```

## write to file

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "tables/metabo_confounder_cors.txt")
write.table(metabo_confounder_cors, file = f, 
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

# metabo_confounder_cors = read.table(file = f, header = TRUE, as.is = TRUE, sep = "\t")

```




```{r confounder_raster_met, fig.width = 20, fig.height = 8}
### Plot colors
pcol =  wesanderson::wes_palette("Darjeeling2", 5 , type = "discrete")

## define plot data
pdata = metabo_confounder_cors


## BF correction on Pvalues, those not sig turn to 0
w = which(pdata$pvalue > 0.05/40)
pdata$pvalue[w] = 0


## Define Covariate plot order
x = pdata %>% group_by(covariate) %>% summarize(mean_etasq = mean(etasq) ) %>% arrange(desc(mean_etasq) )
pdata$covariate = factor(pdata$covariate, levels = x$covariate)


###
PlotA = pdata %>% ggplot( aes(y = covariate, x = dependent) ) +
  geom_tile(aes(fill = etasq)) +
  scale_fill_gradient(low = "white" , high = pcol[2] ) +
  theme(axis.text.x = element_text(angle = 90, size = 3)) +
  geom_point(aes(color = -log10(pvalue) , size = -log10(pvalue) ) , shape = 19 ) +
  scale_color_gradient(low = "pink", high = "red") + 
  labs(color = "-log10(P)", size = "-log10(P)") +
  theme(legend.position="bottom")

PlotA


```


## PLOT PDF TO FILE

```{r}
w = grep("results_dir", par_data[,1])
f = paste0(par_data[w,2], "figures/PGS_Metabolite_confounder_EtaSq.pdf")
pdf(f, width = 15, height = 8)
PlotA
dev.off()


```



