---
title: "NEO BMI-Metabolite point estimate derivation"
author: "David Hughes"
date: "2/6/2021"
output: 
  pdf_document:
    number_sections: true
---

This script and markdown will derive each of the following estimates:

    1. observational point estimates    
        - weighted NEO
        - Leiderdorp
        - NEO
        - Leiden

    2. one-sample MR point estimates
        - weighted NEO
        - Leiderdorp
        - NEO
        - Leiden


```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)

source("LOAD.R")

library(ivreg)
```



## Weighted NEO analysis

```{r wNEO}
wNEO_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_stats = as.data.frame(wNEO_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_stats[,i] = as.numeric(wNEO_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_wNEO_estimates.txt")
# write.table(wNEO_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
wNEO_long = obsmr_2_longformat(data = wNEO_stats, 
                               pop_id = "wNEO", 
                               alpha = 0.05/43, 
                               sandwich_estimates = FALSE)
```

## Those metabolites causally influenced by BMI

```{r}
wNEO_stats %>% 
  filter(MR_P < 0.05/43) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```

## response amino acids

```{r}
wNEO_stats %>% 
  filter(MR_outcome %in% c("ala_r","leu_r","ile_r","val_r","his_r","phe_r","tyr_r","gln_r", 
                           "ala_f","leu_f","ile_f","val_f","his_f","phe_f","tyr_f","gln_f", 
                           "ala_p","leu_p","ile_p","val_p","his_p","phe_p","tyr_p","gln_p") ) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```



## Weighted NEO female

```{r Leiderdorp}
female = which( mydata$sex == "female")
##
wNEO_female_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[female, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_female_stats = as.data.frame(wNEO_female_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_female_stats[,i] = as.numeric(wNEO_female_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_wNEO_females_estimates.txt")
write.table(wNEO_female_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
wNEO_female_long = obsmr_2_longformat(data = wNEO_female_stats, 
                                     pop_id = "wNEO_female", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```


## Weighted NEO male

```{r Leiderdorp}
male = which( mydata$sex == "male")
##
wNEO_male_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[male, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_male_stats = as.data.frame(wNEO_male_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_male_stats[,i] = as.numeric(wNEO_male_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_wNEO_males_estimates.txt")
write.table(wNEO_male_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
wNEO_male_long = obsmr_2_longformat(data = wNEO_male_stats, 
                                     pop_id = "wNEO_male", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```


## NEO

```{r NEO}
NEO_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex"),
            weights = NA,
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
NEO_stats = as.data.frame(NEO_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  NEO_stats[,i] = as.numeric(NEO_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_NEO_estimates.txt")
write.table(NEO_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
NEO_long = obsmr_2_longformat(data = NEO_stats, 
                              pop_id = "NEO", 
                              alpha = 0.05/43, 
                              sandwich_estimates = FALSE)
```



## Leiderdorp

```{r Leiderdorp}
Leiderdorp = grep("Leiderdorp", mydata$subpop)
##
Leiderdorp_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[Leiderdorp, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","age","sex"),
            weights = NA,
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
Leiderdorp_stats = as.data.frame(Leiderdorp_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  Leiderdorp_stats[,i] = as.numeric(Leiderdorp_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_Leiderdorp_estimates.txt")
write.table(Leiderdorp_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
Leiderdorp_long = obsmr_2_longformat(data = Leiderdorp_stats, 
                                     pop_id = "Leiderdorp", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```



## Leiderdorp female

```{r Leiderdorp}
Leiderdorp_female = which( mydata$subpop == "Leiderdorp" & mydata$sex == "female")
##
Leiderdorp_female_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[Leiderdorp_female, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","age"),
            weights = NA,
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
Leiderdorp_female_stats = as.data.frame(Leiderdorp_female_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  Leiderdorp_female_stats[,i] = as.numeric(Leiderdorp_female_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_Leiderdorp_females_estimates.txt")
write.table(Leiderdorp_female_stats, file = f, 
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
Leiderdorp_female_long = obsmr_2_longformat(data = Leiderdorp_female_stats, 
                                     pop_id = "Leiderdorp_female", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```


## Leiderdorp male

```{r Leiderdorp}
Leiderdorp_male = which( mydata$subpop == "Leiderdorp" & mydata$sex == "male")
##
Leiderdorp_male_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[Leiderdorp_male, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","age"),
            weights = NA,
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
Leiderdorp_male_stats = as.data.frame(Leiderdorp_male_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  Leiderdorp_male_stats[,i] = as.numeric(Leiderdorp_male_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_Leiderdorp_males_estimates.txt")
write.table(Leiderdorp_male_stats, file = f, 
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
Leiderdorp_male_long = obsmr_2_longformat(data = Leiderdorp_male_stats, 
                                     pop_id = "Leiderdorp_male", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```




## Leiden

```{r Leiden}
Leiden = grep("Leiden", mydata$subpop)
##
Leiden_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata[Leiden, ],
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","age","sex"),
            weights = NA,
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
Leiden_stats = as.data.frame(Leiden_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  Leiden_stats[,i] = as.numeric(Leiden_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_Leiden_estimates.txt")
write.table(Leiden_stats, file = f, 
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
Leiden_long = obsmr_2_longformat(data = Leiden_stats, 
                                     pop_id = "Leiden", 
                                     alpha = 0.05/43, 
                                     sandwich_estimates = FALSE)
```



## Save the data to Rdata file

```{r}
obs_tsls_est = list(wNEO_stats, wNEO_female_stats, wNEO_male_stats,
                    Leiderdorp_stats, Leiderdorp_female_stats,
                    Leiderdorp_male_stats,
                    NEO_stats, Leiden_stats)
names(obs_tsls_est) = c("wNEO", "wNEO_female", "wNEO_male",
                        "Leiderdorp", "Leiderdorp_female","Leiderdorp_male", 
                        "NEO","Leiden")
####
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]

f = paste0(resultsdir, "obs_tsls_estimates_Dec12_2022.Rdata")
save( obs_tsls_est, file = f)
```


## Save the data to Rdata file

```{r}
obs_tsls_est_long = list(wNEO_long, wNEO_female_long, wNEO_male_long,
                    Leiderdorp_long, Leiderdorp_female_long,
                    Leiderdorp_male_long,
                    NEO_long, Leiden_long)
names(obs_tsls_est_long) = c("wNEO", "wNEO_female", "wNEO_male",
                        "Leiderdorp", "Leiderdorp_female","Leiderdorp_male", 
                        "NEO","Leiden")
####
f = paste0(resultsdir, "obs_tsls_estimates_long_Dec12_2022.Rdata")
save( obs_tsls_est_long, file = f)

```


# SENSITIVITY ANALYSIS

## extract the perox3 samples

```{r}
w = which(mydata$perox_3 == 1)
tempdata = mydata[-w, ]
```

## Weighted NEO analysis

```{r wNEO}
wNEO_sensitivity = t( sapply(traits, function(trait){
  out = obsmr( wdata = tempdata,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex", "PC3", "packyears","diet_weigth_loss"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_sensitivity = as.data.frame(wNEO_sensitivity)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_sensitivity[,i] = as.numeric(wNEO_sensitivity[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Dec12_2022_wNEO_SENSITIVITY_estimates.txt")
write.table(wNEO_sensitivity, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

##########################
## convert 2 long format
##########################
wNEO_sensitivity_long = obsmr_2_longformat(data = wNEO_sensitivity, 
                               pop_id = "wNEO_SA", 
                               alpha = 0.05/43, 
                               sandwich_estimates = FALSE)
```



## Weighted NEO analysis with NO rank normal transformation

```{r wNEO}
wNEO_noRNT_stats = t( sapply(traits, function(trait){
  out = obsmr( wdata = mydata,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex"),
            weights = "pweight",
            rnt_outcome = FALSE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_noRNT_stats = as.data.frame(wNEO_noRNT_stats)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_noRNT_stats[,i] = as.numeric(wNEO_noRNT_stats[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/Mar7_2023_wNEO_noRNT_estimates.txt")
write.table(wNEO_noRNT_stats, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

```

## Those metabolites causally influenced by BMI

```{r}
wNEO_noRNT_stats %>% 
  filter(MR_P < 0.05/43) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```


# Alternative response traits 

  1. ornls
  2. delta

## Setup the data

```{r}
## find the response traits
w = grep("_r", colnames(study_data$full_data) )
ornls = study_data$full_data[,w]
## remove the deming response traits
w = grep("_deming_r", colnames(ornls) )
ornls = ornls[,-w]
## rename traits
colnames(ornls) = gsub("_r","_ornls",colnames(ornls))

## derive delta
delta = sapply(metids, function(id){
  f = paste0(id,"_f")
  p = paste0(id,"_p")
  d = mydata[,p] - mydata[,f]
  return(d)
}) 
colnames(delta) = paste0(metids, "_delta")

alt_res_data = cbind(mydata[,1:11] , ornls, delta)
alr_res_traits = c( colnames(ornls), colnames(delta) )

```


## Weighted NEO analysis on ornls and delta response traits

```{r wNEO}
wNEO_alt_response = t( sapply(alr_res_traits, function(trait){
  out = obsmr( wdata = alt_res_data,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 10,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_alt_response = as.data.frame(wNEO_alt_response)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_alt_response[,i] = as.numeric(wNEO_alt_response[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/May17_wNEO_alt_response_estimates.txt")
write.table(wNEO_alt_response, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

```

## Those metabolites causally influenced by BMI

```{r}
wNEO_alt_response %>% 
  filter(MR_P < 0.05/43) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```


```{r}
wNEO_alt_response %>% 
  filter(MR_P < 0.05) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  arrange(MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```


## Weighted NEO analysis on ornls and delta response traits

```{r wNEO}
wNEO_alt_response_v2 = t( sapply(alr_res_traits, function(trait){
  out = obsmr( wdata = alt_res_data,
            outcome = trait,
            exposure = "bmi",
            instrument = "Yengo_Weighted_GRS",
            covariates = c("visit_date","subpop","age","sex"),
            weights = "pweight",
            rnt_outcome = TRUE,
            outlier_method = "iqr",
            outlier_cutoff = 5,
            messages = TRUE)
  return(out)
}) )
##########################
## turn into a data.frame
##########################
wNEO_alt_response_v2 = as.data.frame(wNEO_alt_response_v2)
##########################
## convert to numeric
##########################
num_cols = c(4:6, 8:9, 12:24, 27:36, 39:48, 51:74 )
for(i in num_cols){
  wNEO_alt_response_v2[,i] = as.numeric(wNEO_alt_response_v2[,i])
}
##########################
## write table to file
##########################
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
f = paste0(resultsdir, "tables/May17_wNEO_alt_response_estimates_v2.txt")
write.table(wNEO_alt_response_v2, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

```




```{r}
wNEO_alt_response_v2 %>% 
  filter(MR_P < 0.05/43) %>% 
  dplyr::select(MR_outcome, MR_n, MR_beta, MR_se, MR_P) %>%
  arrange(MR_P) %>%
  kable() %>% kableExtra::kable_classic()
```


