---
title: "NEO BMI-Metabolite Exploring Obs Point Estimate Results"
author: "David Hughes"
date: "13/12/2022"
output: 
  pdf_document:
    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)

source("LOAD.R")

```


## Basic Summary Statistics for each metabolite trait

```{r}
trait_sum_stats = sapply(traits, function(trait){
  x = na.omit( mydata[, trait] )
  ###
  n = length( na.omit(x))
  a = mean(x, na.rm = TRUE)
  # b = median(x, na.rm = TRUE)
  d = sd(x, na.rm = TRUE)
  e = var(x, na.rm = TRUE)
  q = quantile(x, probs = c(0, 0.025, 0.25, 0.5, 0.75, 0.975, 1) , na.rm = TRUE)
  s = psych::skew(x)
  k = psych::kurtosi(x)
  if(length(x)>5000){
    x = sample(x, 5000)
  }
  w = shapiro.test(x)$statistic
  ##
  out = c(n, a, d, e, s, k, w, q)
  names(out) = c("N", "mean", "sd", "var", 
                 "skew", "kurtosis",
                 "shapiro_W",
                 "min", 
                 "quantile_0.025",
                 "quantile_0.25",
                 "quantile_0.5",
                 "quantile_0.75",
                 "quantile_0.975", 
                 "max")
  return(out)
})

trait_sum_stats = as.data.frame(t(trait_sum_stats))
  
```


## add to feature annotation

```{r}
fdata_out = cbind(study_data$feature_anno, trait_sum_stats)
fdata_out[1:5,] %>% kable() %>% kableExtra::kable_classic()
```

## Write feature annotation data to flat text file

```{r}
w = which(par_data[,1] == "results_dir")
resultsdir = par_data[w,2]
###
f = paste0(resultsdir, "tables/feature_annotation_sum_stats.txt")
write.table(fdata_out, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)
```


## Mean and quantiles for each metabolite trait by BMI class

```{r}
## Define BMI cats levels
bmicats = levels(mydata$bmi_cat)
### an empty element to store the data
bmicat_trait_sum_stats = c()
###
for(cat in bmicats){
  for(trait in traits){
    # BMI cat subset
    a = which(mydata$bmi_cat == cat)
    ## Dietary State of trait
    ds = strsplit(trait, split="_")[[1]][2]
    if(ds == "f"){ds = "fasting"}
    if(ds == "p"){ds = "postprandial"}
    if(ds == "r"){ds = "response"}
    # N, mean, quantiles
    n = length( na.omit( mydata[a,trait] ) )
    m = mean( mydata[a,trait] , na.rm = TRUE)
    q = quantile( mydata[a,trait] , probs = c(0.025, 0.25, 0.5, 0.75, 0.975) , na.rm = TRUE)
    ## data out
    out = c(trait, ds, cat, n, m, q)
     names(out) = c("trait","dietary_state","BMI_category",
       "N", "mean", 
       "quantile_0.025",
       "quantile_0.25",
       "quantile_0.5",
       "quantile_0.75",
       "quantile_0.975"
                 )
     ### add data to data frame
     bmicat_trait_sum_stats = rbind(bmicat_trait_sum_stats, out)
  }
}
## Define as a data frame
bmicat_trait_sum_stats = as.data.frame(bmicat_trait_sum_stats)
for(i in 4:10){bmicat_trait_sum_stats[,i] = as.numeric(bmicat_trait_sum_stats[,i])}
## Define levels for BMI category
bmicat_trait_sum_stats$BMI_category = factor(bmicat_trait_sum_stats$BMI_category, levels = bmicats)
  
```


## add annotation to the bmicat_trait_sum_stats data frames

```{r}
m = match(bmicat_trait_sum_stats$trait, study_data$feature_anno$ids)
bmicat_trait_sum_stats = cbind(bmicat_trait_sum_stats, 
                               study_data$feature_anno[, c("metabolite","class","subclass", "label", "label.no.units") ])

  
```



## Define vector of Lipoproteins

```{r}
lipoproteins = c("xxlvldl","xlvldl","lvldl","mvldl","svldl", "xsvldl", ## VLDL
      "idl",                                                  ## IDL
      "lldl","mldl","sldl",                                   ## LDL
      "xlhdl","lhdl","mhdl", "shdl")                          ## HDL
```

## Define vector of Lipids suffixes

```{r}
lipids = c("l","pl","c","ce", "fc","tg")
names(lipids) = c("total lipids","phospholipids","cholesterol","cholesterol esters","free cholesterol","triglycerides")
```

## Define vector of Lipid Traits

```{r}
lipid_traits = c()
for(i in lipoproteins){
  for(j in lipids){
    o = paste0(i, j)
    o = c(i, j, o)
    lipid_traits = rbind(lipid_traits, o)
  }
}
## redefine the lipid
m = match(lipid_traits[,2], lipids)
lipid_traits[,2] = names(lipids)[m]

colnames(lipid_traits) = c("lipoprotein","lipid","metabolite")
lipid_traits = as.data.frame(lipid_traits)
```


## Fasting Lipoprotien Summary Stats 

```{r}
## Make row names a column value
trait_sum_stats$traits = rownames(trait_sum_stats)
## fasting values
k = sapply(lipid_traits[,3], function(x){ which(  trait_sum_stats$traits == paste0(x, "_f"))})
fasting_lipoprotein_ss = trait_sum_stats[k, c("traits", "mean", "quantile_0.25", "quantile_0.75")]


## Define metabolite
fasting_lipoprotein_ss$metabolites = sapply(fasting_lipoprotein_ss$traits, function(x){
  strsplit(x, split = "_f")[[1]][1]
})

## Define lipid and lipoprotein
m = match(fasting_lipoprotein_ss$metabolite, lipid_traits$metabolite)
fasting_lipoprotein_ss$lipid = lipid_traits$lipid[m]
fasting_lipoprotein_ss$lipoprotein = lipid_traits$lipoprotein[m]

## Define factor levels for lipids and lipoproteins
fasting_lipoprotein_ss$lipoprotein = factor(fasting_lipoprotein_ss$lipoprotein, 
                                            levels = lipoproteins )
fasting_lipoprotein_ss$lipid = factor(fasting_lipoprotein_ss$lipid, 
                                      levels = c("total lipids","cholesterol",
                                                 "cholesterol esters",
                                                 "phospholipids","free cholesterol","triglycerides"))

fasting_lipoprotein_ss$dietary_state = "fasting"
```


## Postprandial Lipoprotien Summary Stats 

```{r}
## Make row names a column value
trait_sum_stats$traits = rownames(trait_sum_stats)
## fasting values
k = sapply(lipid_traits[,3], function(x){ which(  trait_sum_stats$traits == paste0(x, "_p"))})
postprandial_lipoprotein_ss = trait_sum_stats[k, c("traits", "mean", "quantile_0.25", "quantile_0.75")]


## Define metabolite
postprandial_lipoprotein_ss$metabolites = sapply(postprandial_lipoprotein_ss$traits, function(x){
  strsplit(x, split = "_p")[[1]][1]
})

## Define lipid and lipoprotein
m = match(postprandial_lipoprotein_ss$metabolite, lipid_traits$metabolite)
postprandial_lipoprotein_ss$lipid = lipid_traits$lipid[m]
postprandial_lipoprotein_ss$lipoprotein = lipid_traits$lipoprotein[m]

## Define factor levels for lipids and lipoproteins
postprandial_lipoprotein_ss$lipoprotein = factor(postprandial_lipoprotein_ss$lipoprotein, 
                                            levels = lipoproteins )
postprandial_lipoprotein_ss$lipid = factor(postprandial_lipoprotein_ss$lipid, 
                                      levels = c("total lipids","cholesterol",
                                                 "cholesterol esters",
                                                 "phospholipids","free cholesterol","triglycerides"))

postprandial_lipoprotein_ss$dietary_state = "postprandial"
```




```{r, fig.width = 12, fig.height = 5}
lipoprotein_ss = rbind(fasting_lipoprotein_ss, postprandial_lipoprotein_ss)

( p = lipoprotein_ss %>% ggplot(aes(x = lipoprotein, y = mean )) +
  geom_point(aes(color = lipid), size = 3, alpha = 0.6) +
  geom_line(aes(color = lipid, group = lipid)) +
  theme_bw() +
  labs(y = "mean [mmol/l]") +
  facet_wrap( .~dietary_state ,nrow = 1) +
  theme(legend.position="bottom") )
```



## save fasting lipoproteins beta profile

```{r}
f = paste0(resultsdir, "figures/mean_lipoprotein_concentrations.pdf")
pdf(file = f, width = 12, height = 5)
p
dev.off()
```


## A T-test for difference between fasting and non-fasting abundance

```{r}
f_p_ttest = t( sapply(metids, function(id){
  f = paste0(id, "_f"); f = mydata[, f]
  p = paste0(id, "_p"); p = mydata[, p]
  ###########################
  ### SCALE AND CENTER DATA
  ###########################
  l = length(f)
  d = scale(c(f,p))[,1]
  f = d[1:l]
  p = d[c(l+1):length(d)]
  ##
  x = t.test( p,f, paired = TRUE)
  
  ## T-stat and Pval
  tstat = c(x$alternative, x$statistic, x$parameter, x$p.value)
  names(tstat) = c("test","Tstat","df","P")
  ## mean diff and 95CI
  mdif = c(x$estimate, x$conf.int[1:2])
  names(mdif) = c("mean_dif","L_95CI","U_95CI")
  ## data out
  out = c(tstat, mdif)
  return(out)
}) )
## define as data frame
f_p_ttest = as.data.frame(f_p_ttest)
## set numerics
for(i in 2:7){ f_p_ttest[,i] = as.numeric(f_p_ttest[,i]) }
```


```{r}
f = study_data$feature_anno %>% filter(dietary_state == "fasting")
m = match(rownames(f_p_ttest), f$metabolite)
f_p_ttest = cbind(f_p_ttest, f[m,c("metabolite","raw.label","class","subclass","label","label.no.units")])
```

```{r}
f = paste0(resultsdir, "tables/fasting_postprandial_t_test_scaled_centred.txt")
write.table(f_p_ttest, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)
```



## How many metabolites are different between the two dietary states ?

```{r}
cat(paste0("How many metabolites are different between the two dietary states? \t"))
sum( f_p_ttest$P < 0.05/nrow(f_p_ttest) )

cat(paste0("How many metabolites are NOT different between the two dietary states? \t"))
sum( f_p_ttest$P > 0.05/nrow(f_p_ttest) )

```


## What are the 14 metabolites that do NOT differ between fasting and postprandial states?

```{r}
f_p_ttest %>% filter(P > 0.05/nrow(f_p_ttest) ) %>%
  kable() %>% kableExtra::kable_classic()
```

## Direction of change

```{r}
f_p_ttest %>% filter(P< 0.05/229) %>% summarize( pos_count = sum( mean_dif > 0),
                                                 neg_count = sum( mean_dif < 0) )
```



```{r}
f_p_ttest %>% filter(P< 0.05/229) %>% 
  group_by(class) %>%
  summarize( pos_count = sum( mean_dif > 0),
             neg_count = sum( mean_dif < 0) ) %>% 
  kable() %>% kableExtra::kable_classic()
```

## organize by largest mean dif

```{r}
f_p_ttest %>% arrange(desc(mean_dif)) %>%
  kable() %>% kableExtra::kable_classic()
```


## organize by smallest mean dif

```{r}
f_p_ttest %>% arrange( mean_dif ) %>%
  kable() %>% kableExtra::kable_classic()
```

## Delta distributions

```{r}
delta_ss = t( sapply(metids, function(id){
  f = paste0(id, "_f"); f = mydata[, f]
  p = paste0(id, "_p"); p = mydata[, p]
  ##
  l = length(f)
  x = scale(c(f,p))
  f = x[1:l]
  p = x[(l+1):length(x)]
  ##
  delta = p - f
  m = mean(delta, na.rm = TRUE)
  q = quantile(delta, probs = c(0, 0.025, 0.25, 0.75, 0.975, 1), na.rm = TRUE )
  ss = c(id, m,q)
  names(ss) = c("metabolite","mean","min","quantile_2.5","quantile_25","quantile_75","quantile_97.5","max")
  return(ss)
}) )

### define as data frame
delta_ss = as.data.frame(delta_ss)
for(i in 2:ncol(delta_ss)){ delta_ss[,i] = as.numeric(delta_ss[,i]) }

### add annotation
m = match(delta_ss$metabolite, study_data$feature_anno$metabolite)
delta_ss = cbind(delta_ss, study_data$feature_anno[m, c("class","subclass")] )

## define metabolite factor levels
delta_ss$metabolite = factor(delta_ss$metabolite, levels = delta_ss$metabolite)
```


```{r, fig.width = 15, fig.height = 5}
pcol = c( RColorBrewer::brewer.pal(6, "Set1"), RColorBrewer::brewer.pal(8, "Dark2") )[14:1]


delta_ss %>% ggplot(aes(x = metabolite, y = mean)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 1.5) +
  geom_point(aes(color = class) ,alpha = 0.6, size = 1.5) +
  scale_color_manual(values = pcol) + 
  geom_errorbar(aes(ymin = quantile_2.5, ymax = quantile_97.5, color = class), linewidth = 0.5) +
  theme_bw() +
  labs(y = "delta: postprandial - fasting", x = "metabolite") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 6)) +
  theme(legend.position = "bottom")

```



```{r, fig.width = 15, fig.height = 8}
pcol = c( RColorBrewer::brewer.pal(6, "Set1"), RColorBrewer::brewer.pal(8, "Dark2") )[14:1]

k = c("Lipoprotein subclasses", "Lipoprotein subclasses ratios")

plot1 = delta_ss %>% 
  filter( !class %in% k ) %>% 
  ggplot(aes(x = metabolite, y = mean)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1.25) +
  geom_point(aes(color = class) ,alpha = 0.6, size = 1.5) +
  scale_color_manual(values = pcol) + 
  geom_errorbar(aes(ymin = quantile_2.5, ymax = quantile_97.5, color = class), linewidth = 0.5) +
  theme_bw() +
  labs(y = "delta: postprandial - fasting", x = "metabolite") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)) +
  theme(legend.position = "right") +
  facet_wrap(.~class, nrow = 1, scales = "free_x") + theme(strip.text.x = element_blank())

##########
temp = delta_ss %>% filter( class %in% k )
temp$subclass = gsub(" ratios","", as.character( temp$subclass ) )
temp$subclass = factor(temp$subclass, 
                       levels = c("Extremely large VLDL","Very large VLDL", 
                                  "Large VLDL","Medium VLDL","Small VLDL","Very Small VLDL",
                                  "IDL",
                                  "Large LDL","Medium LDL","Small LDL",
                                  "Very large HDL","Large HDL","Medium HDL","Small HDL") )
#
plot2 = temp %>% 
  ggplot(aes(x = metabolite, y = mean)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1.25) +
  geom_point(aes(color = subclass) ,alpha = 0.6, size = 1.5) +
  scale_color_manual(values = pcol) + 
  geom_errorbar(aes(ymin = quantile_2.5, ymax = quantile_97.5, color = subclass), linewidth = 0.5) +
  theme_bw() +
  labs(y = "delta: postprandial - fasting", x = "metabolite") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)) +
  theme(legend.position = "right") +
  facet_wrap(.~class, ncol = 1, scales = "free_x") 

```


```{r, fig.width = 15, fig.height = 12}
library(patchwork)
plot1 / plot2 + plot_layout(heights = c(1.3,3))

```

## save delta plot

```{r}
f = paste0(resultsdir, "figures/scaled_delta_mean_95CI.pdf")
pdf(file = f, width = 12, height = 13)
plot1 / plot2 + plot_layout(heights = c(1.3,3))
dev.off()
```


## Are any metabolites uniformly positive or negative in their delta value across samples?

```{r}
delta_ss %>% filter(quantile_2.5 > 0 & quantile_97.5 > 0)
delta_ss %>% filter(quantile_2.5 < 0 & quantile_97.5 < 0)
```

## Do the simple Delta and deming regression residuals correlated ?

```{r}
delta_response_cors = sapply(metids, function(id){
  f = paste0(id, "_f")
  p = paste0(id, "_p")
  r = paste0(id, "_r")
  delta = mydata[, p] - mydata[,f]
  response = mydata[,r]
  a = cor.test(delta, response)
  return(a$estimate)
})
mean(delta_response_cors)
quantile(delta_response_cors, probs = c(0.025, 0.975))
```









## Total lipid concentrations by BMI category





```{r, fig.width = 10, fig.height = 17}
## total lipids
Ltraits = paste0(lipoproteins, "l_f")
## make the plot
p1 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "total lipid (mmol/L)",
                           scales = "fixed")

## cholesterol
Ltraits = paste0(lipoproteins, "c_f")
## make the plot
p2 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "cholesterol (mmol/L)")

## phospholipids
Ltraits = paste0(lipoproteins, "pl_f")
## make the plot
p3 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "phospholipids (mmol/L)")

## cholesterol esters
Ltraits = paste0(lipoproteins, "ce_f")
## make the plot
p4 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "cholesterol esters (mmol/L)")

## triglycerides
Ltraits = paste0(lipoproteins, "tg_f")
## make the plot
p5 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "triglycerides (mmol/L)")

## free cholesterol
Ltraits = paste0(lipoproteins, "fc_f")
## make the plot
p6 = lipid_conc_plot_by_bmi_cat(wdata = mydata, 
                           trait_vector = Ltraits, 
                           ylabel = "free cholesterol (mmol/L)")


# ggpubr::ggarrange(p1, p2, ncol = 1)  

library(patchwork)
p1 / p2 / p3 / p4 / p5 / p6 + plot_layout(guides = 'collect') & theme(legend.position = 'bottom')
```



